Global adjoint tomography -- Model GLAD-M25\footnote{This paper is dedicated to Dimitri Komatitsch, computational geophysicist extraordinaire, who passed away unexpectedly shortly before its submission.}

### 2019-02-11 main.tex
Updates from Overleaf

The observed seismograms need to be compared to corresponding synthetic seismograms
for the purpose of making measurements.
The forward simulations are based on the global spectral-element solver
SPECFEM3D\_GLOBE~\citep{KoTr02a,KoTr02a}, and take a CMT solution, a relevant list
of stations, and the current earth model as input.
The solver is GPU accelerated, and the simulations are performed on the Cray supercomputer ``Titan''
at the OLCF.
To simulate 120~min seismograms with a shortest period of 17~s on 384 NVIDIA
Tesla K20X GPUs takes approximately 10~min of wall clock time. {\color{red} Wenjie: is that correct?}

@@

When we compare GLAD-M25 with other global models,
we find overall good agreement at longer wavelengths.
However, unlike other global models, when we compare GLAD-M25 with regional models
we also observe generally good agreement at smaller wavelengths,
and, in particular, plumes and subduction zones feature prominently.

Obvious future directions of research include growing the earthquake database further to the roughly 6,000 earthquakes in the right magnitude range already available via the Global CMT database
and reducing the shortest simulation period further from 17~s to 8.5~s.
The most recent machine at OLCF, called ``Summit'', uses IBM Power9 CPUs accelerated by NVIDIA Tesla V100 GPU and can readily accommodate such simulations.

@@

\begin{acknowledgments}
This research used resources of the Oak Ridge Leadership Computing Facility,
which is a DOE Office of Science User Facility supported under contract DE-AC05-00OR22725.
Additional computational resources were provided by the Princeton Institute
for Computational Science \& Engineering (PICSciE).
We acknowledge IRIS ({\tt iris.edu}) and ORFEUS ({\tt orfeus-eu.org}) for
providing the data used in this study.
We thank Ryan Modrak, Ridvan \"{O}rsvuran, Frederik J. Simons, and James Smith for fruitful discussions,
and Caio Ciardelli for implementing the spherical harmonic model expansion.
The open source spectral-element software package SPECFEM3D\_GLOBE and
the seismic measurement software package FLEXWIN used for this article are
 freely available via the Computational Infrastructure for Geodynamics
 (CIG; {\tt geodynamics.org}). This research was supported by NSF grant~1644826.
\end{acknowledgments}

@@

Instead of the complex spherical harmonic model expansion~(\ref{eq:m}),
we may wish to use a real spherical harmonic expansion of the form~\citep[][Section~B.8, Eqn.~B.99]{DT98}
\begin{equation}
\begin{split}
    m(\mathbf{x}) \ = & \ \sum_{n=0}^N\sum_{\ell = 0}^L \left[ {}_na_{\ell 0}\,R_n(r)\,X_{\ell 0}(\theta)
   +\sqrt{2}\sum_{m=1}^\ell R_n(r)\,({}_na_{\ell m}\,\cos m\phi+{}_nb_{\ell m}\,\sin m\phi)\,X_{\ell m}(\theta)\right]
    \quad ,
\end{split}
    \label{eq:mreal}
\end{equation}
where~\citep[][Eqn.~B.30]{DT98}
\begin{equation}
    Y_{\ell m}(\theta,\phi)=X_{\ell m}(\theta)\exp(i m\phi)
    \quad ,
\end{equation}
and where
\begin{equation}
    {}_na_{\ell 0}=\sum_{n'}A^{-1}_{nn'}\sum_{\mathrm{elem}}\sum_{\alpha,\beta,\gamma}\omega_\alpha\,\omega_\beta\,\omega_\gamma\,m^{\alpha\beta\gamma}\,J^{\alpha\beta\gamma}\,R_{n'}^{\alpha\beta\gamma}\,X_{\ell m}^{\alpha\beta\gamma}
    \quad ,
\end{equation}
and for~$1\le m\le \ell$
\begin{equation}
    {}_na_{\ell m}=\sqrt{2}\,\sum_{n'}A^{-1}_{nn'}\sum_{\mathrm{elem}}\sum_{\alpha,\beta,\gamma}\omega_\alpha\,\omega_\beta\,\omega_\gamma\,m^{\alpha\beta\gamma}\,J^{\alpha\beta\gamma}\,R_{n'}^{\alpha\beta\gamma}\,X_{\ell m}^{\alpha\beta\gamma}\,(\cos m \phi)^{\alpha\beta\gamma}
    \quad ,
\end{equation}
\begin{equation}
    {}_nb_{\ell m}=\sqrt{2}\,\sum_{n'}A^{-1}_{nn'}\sum_{\mathrm{elem}}\sum_{\alpha,\beta,\gamma}\omega_\alpha\,\omega_\beta\,\omega_\gamma\,m^{\alpha\beta\gamma}\,J^{\alpha\beta\gamma}\,R_{n'}^{\alpha\beta\gamma}\,X_{\ell m}^{\alpha\beta\gamma}\,(\sin m \phi)^{\alpha\beta\gamma}
    \quad .
\end{equation}
The normalized power per degree~$\ell$ may then be calculated via
\begin{equation}
    \sigma_\ell^2=\frac{1}{2\ell+1}\,\sum_{n'=0}^N\,\sum_{n=0}^N\,A_{n'n}\,\left[{}_{n'}a_{\ell 0}\,{}_{n}a_{\ell 0}+\sum_{m=1}^\ell({}_{n'}a_{\ell m}\,{}_na_{\ell m}+{}_{n'}b_{\ell m}\,{}_nb_{\ell m})\right]
    \quad ,
    \label{eq:powerdegree}
\end{equation}
and the normalized power per degree as a function of radius is determined via
\begin{equation}
    \Sigma_\ell^2(r)=\frac{1}{2\ell+1}\,\sum_{n'=0}^N\,\sum_{n=0}^N\,R_{n'}(r)\,R_{n}(r)\,\left[{}_{n'}a_{\ell 0}\,{}_{n}a_{\ell 0}+\sum_{m=1}^\ell({}_{n'}a_{\ell m}\,{}_na_{\ell m}+{}_{n'}b_{\ell m}\,{}_nb_{\ell m})\right]
    \quad .
    \label{eq:powerdegreeradius}
\end{equation}
In Fig.~\ref{fig:powerspectra} we plot the normalized power per degree~$\ell$ for models S362ANI, GLAD-M15, and GLAD-M25,
and in Fig.~\ref{fig:powerspectraradius} we plot the normalized power per degree as a function of radius
for the same set of models.

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{figures/power.pdf}
  \caption{\small{Normalized power per degree~$\ell$, based on Eqn.~(\ref{eq:powerdegree}), for models S362ANI (red), GLAD-M15 (green), and GLAD-M25 (blue).
  Model S362ANI has no power beyond degree~18.
  (Courtesy of Caio Ciardelli.)
  }}
  \label{fig:powerspectra}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.95\textwidth]{figures/power_radius.pdf}
  \caption{\small{Normalized power per degree~$\ell$ as a function of radius, based on Eqn.~(\ref{eq:powerdegreeradius}), for models S362ANI (left), GLAD-M15 (middle), and GLAD-M25 (right).
  (Courtesy of Caio Ciardelli.)
  }}
  \label{fig:powerspectraradius}
\end{figure}

% To assess the accuracy of the expansion,
% we may calculate the model norm in the GLL basis,
% \begin{equation}
%     ||m_\text{\,GLL}||^2=\int_V m_\text{\,GLL}^2(\mathbf{x})\,\mathrm{d}^3\mathbf{x}=
%     \sum_{\mathrm{elem}}\sum_{\alpha,\beta,\gamma}\omega_\alpha\,\omega_\beta\,\omega_\gamma\,(m^{\alpha\beta\gamma})^2\,J^{\alpha\beta\gamma}
%     \quad ,
% \end{equation}
% and compare it to the model norm in the spherical harmonic basis,
% \begin{equation}
%     ||m_\text{\,SH}||^2=\int_V m_\text{\,SH}^2(\mathbf{x})\,\mathrm{d}^3\mathbf{x}=
%     \sum_{\ell = 0}^L(2\ell+1)\,\sigma_\ell^2
%     \quad ,
% \end{equation}
% by calculating the error
% \begin{equation}
%     \epsilon = \log\,\frac{||m_\text{\,SH}||^2}{||m_\text{\,GLL}||^2}
%     \quad .
% \end{equation}

>>>>

The observed seismograms need to be compared to corresponding synthetic seismograms
for the purpose of making measurements.
The forward simulations are based on the global spectral-element solver
SPECFEM3D\_GLOBE~\citep{KoTr02a,KoTr02a}, and take a CMT solution, a relevant list
of stations, and the current earth model as input.
The solver is GPU accelerated, and the simulations are performed on the Cray supercomputer ``Titan''
at the OLCF.
To simulate 120~min seismograms with a shortest period of 17~s on 384 NVIDIA
Tesla K20X GPUs takes approximately 10~min of wall clock time.

@@

When we compare GLAD-M25 with other global models,
we find overall good agreement at longer wavelengths.
However, unlike other global models, when we compare GLAD-M25 with regional models
we also observe generally good agreement at smaller wavelengths,
and, in particular, plumes and subduction zones feature prominently.
Thus, our global inversion is bridging the gap between global and regional studies,
approaching regional-scale resolution is areas with sufficient data coverage.
We can observe this increase in resolution in GLAD-M25 in the normalized power per spherical harmonic degree as a function of radius, which is displayed in Fig.~\ref{fig:powerspectraradius}.

\begin{figure}
  \centering
  \includegraphics[width=0.95\textwidth]{figures/power_radius2.pdf}
  \caption{\small{Normalized power per degree~$\ell$ as a function of radius, based on Eqn.~(\ref{eq:powerdegreeradius}), for models S362ANI (first column), S40RTS (second column) GLAD-M15 (third column), and GLAD-M25 (fourth column).
  Shown are depths greater than 120~km.
  (Courtesy of Caio Ciardelli.)
  }}
  \label{fig:powerspectraradius}
\end{figure}

Obvious future directions of research include growing the earthquake database further to the roughly 6,000 earthquakes in the right magnitude range already available via the Global CMT database,
and reducing the shortest simulation period further from 17~s to 8.5~s.
The most recent machine at OLCF, called ``Summit'', uses IBM Power9 CPUs accelerated by NVIDIA Tesla V100 GPU and can readily accommodate such simulations.

@@

\begin{acknowledgments}
This research used resources of the Oak Ridge Leadership Computing Facility,
which is a DOE Office of Science User Facility supported under contract DE-AC05-00OR22725.
Additional computational resources were provided by the Princeton Institute
for Computational Science \& Engineering (PICSciE).
We acknowledge IRIS ({\tt iris.edu}) and ORFEUS ({\tt orfeus-eu.org}) for
providing the data used in this study.
We thank Ryan Modrak, Ridvan \"{O}rsvuran, Frederik J.\ Simons, and James Smith for fruitful discussions,
and Caio Ciardelli for implementing the spherical harmonic model expansion.
The open source spectral-element software package SPECFEM3D\_GLOBE and
the seismic measurement software package FLEXWIN used for this article are
 freely available via the Computational Infrastructure for Geodynamics
 (CIG; {\tt geodynamics.org}).
 This research was supported by NSF grant~1644826.
\end{acknowledgments}

@@

Instead of the complex spherical harmonic model expansion~(\ref{eq:m}),
we may wish to use a real spherical harmonic expansion of the form~\citep[][Section~B.8, Eqn.~B.99]{DT98}
\begin{equation}
\begin{split}
    m(\mathbf{x}) \ = & \ \sum_{n=0}^N\sum_{\ell = 0}^L \left[ {}_na_{\ell 0}\,R_n(r)\,X_{\ell 0}(\theta)
   +\sqrt{2}\sum_{m=1}^\ell R_n(r)\,({}_na_{\ell m}\,\cos m\phi+{}_nb_{\ell m}\,\sin m\phi)\,X_{\ell m}(\theta)\right]
    \quad ,
\end{split}
    \label{eq:mreal}
\end{equation}
where~\citep[][Eqn.~B.30]{DT98}
\begin{equation}
    Y_{\ell m}(\theta,\phi)=X_{\ell m}(\theta)\exp(i m\phi)
    \quad ,
\end{equation}
and where
\begin{equation}
    {}_na_{\ell 0}=\sum_{n'}A^{-1}_{nn'}\sum_{\mathrm{elem}}\sum_{\alpha,\beta,\gamma}\omega_\alpha\,\omega_\beta\,\omega_\gamma\,m^{\alpha\beta\gamma}\,J^{\alpha\beta\gamma}\,R_{n'}^{\alpha\beta\gamma}\,X_{\ell m}^{\alpha\beta\gamma}
    \quad ,
\end{equation}
and for~$1\le m\le \ell$
\begin{equation}
    {}_na_{\ell m}=\sqrt{2}\,\sum_{n'}A^{-1}_{nn'}\sum_{\mathrm{elem}}\sum_{\alpha,\beta,\gamma}\omega_\alpha\,\omega_\beta\,\omega_\gamma\,m^{\alpha\beta\gamma}\,J^{\alpha\beta\gamma}\,R_{n'}^{\alpha\beta\gamma}\,X_{\ell m}^{\alpha\beta\gamma}\,(\cos m \phi)^{\alpha\beta\gamma}
    \quad ,
\end{equation}
\begin{equation}
    {}_nb_{\ell m}=\sqrt{2}\,\sum_{n'}A^{-1}_{nn'}\sum_{\mathrm{elem}}\sum_{\alpha,\beta,\gamma}\omega_\alpha\,\omega_\beta\,\omega_\gamma\,m^{\alpha\beta\gamma}\,J^{\alpha\beta\gamma}\,R_{n'}^{\alpha\beta\gamma}\,X_{\ell m}^{\alpha\beta\gamma}\,(\sin m \phi)^{\alpha\beta\gamma}
    \quad .
\end{equation}
The normalized power per degree~$\ell$ may then be calculated via
\begin{equation}
    \sigma_\ell^2=\frac{1}{2\ell+1}\,\sum_{n'=0}^N\,\sum_{n=0}^N\,A_{n'n}\,\left[{}_{n'}a_{\ell 0}\,{}_{n}a_{\ell 0}+\sum_{m=1}^\ell({}_{n'}a_{\ell m}\,{}_na_{\ell m}+{}_{n'}b_{\ell m}\,{}_nb_{\ell m})\right]
    \quad ,
    \label{eq:powerdegree}
\end{equation}
and the normalized power per degree as a function of radius is determined via
\begin{equation}
    \Sigma_\ell^2(r)=\frac{1}{2\ell+1}\,\sum_{n'=0}^N\,\sum_{n=0}^N\,R_{n'}(r)\,R_{n}(r)\,\left[{}_{n'}a_{\ell 0}\,{}_{n}a_{\ell 0}+\sum_{m=1}^\ell({}_{n'}a_{\ell m}\,{}_na_{\ell m}+{}_{n'}b_{\ell m}\,{}_nb_{\ell m})\right]
    \quad .
    \label{eq:powerdegreeradius}
\end{equation}
In Fig.~\ref{fig:powerspectra} we plot the normalized power per degree~$\ell$ for models S362ANI, S40RTS, GLAD-M15, and GLAD-M25,
and Fig.~\ref{fig:powerspectraradius} shows the normalized power per degree as a function of radius
for the same set of models.

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{figures/power2.pdf}
  \caption{\small{Normalized power per degree~$\ell$, based on Eqn.~(\ref{eq:powerdegree}), for models S362ANI (red), S40RTS (orange), GLAD-M15 (green), and GLAD-M25 (blue).
  The calculation includes depths greater than 120~km.
  Model S362ANI has no power beyond degree~18,
  and model S40RTS has no power beyond degree~40.
  (Courtesy of Caio Ciardelli.)
  }}
  \label{fig:powerspectra}
\end{figure}
